Automated residual plot assessment with the \textsf{R} package \textsf{autovi} and the \textsf{Shiny} app \textsf{autovi.web}

### 2025-09-04 autovi_paper.tex
Fix math notaion and some packages

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Automated residual plot assessment with the R package  and the Shiny app },
  pdfauthor={Weihao Li; Dianne Cook; Emi Tanaka; Susan VanderPlas; Klaus Ackermann},
  pdfkeywords={initial data analysis, statistical graphics, data
visualization, visual inference, computer vision, machine
learning, hypothesis testing, regression analysis, model diagnostics},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{Automated residual plot assessment with the R package
\textsf{autovi} and the Shiny app \textsf{autovi.web}}

@@

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{equation}

here, \(p(.)\) and \(q(.)\) are the probability density functions of the
reference residual distribution \(P\) and the true residual distribution
\(Q\), respectively.

This distance measure depends on knowledge of the true residual
distribution, which is unknown in practice. To compute \(D_{KL}\) for
the training samples, Equation \ref{eq:kl-1} takes different forms
depending on the specific model violations. For instance, where
necessary higher-order predictors, \(\boldsymbol{Z}\), and their
corresponding parameter, \(\boldsymbol{\beta}_Z\), are omitted from the
fitted linear model, the distance measure can be expanded as follows:

@@

\section{R package: autovi}\label{sec-autovi}

The main purpose of \texttt{autovi} is to provide rejection decisions
and \(p\)-values for testing the null hypothesis (\(H_0\)) that the
regression model is correctly specified. The package provides automated
interpretation of residual plots using computer vision. The name
\texttt{autovi} stands for \textbf{auto}mated \textbf{v}isual
\textbf{i}nference. This functionality can be accessed through the R
package \texttt{autovi}, or through a web interface,
\texttt{autovi.web}, which enables users to perform analyses without
installing R, Python, or their associated dependencies locally.

@@

The \texttt{autovi} infrastructure effectively integrates multiple
programming languages and libraries into a comprehensive analytical
tool. It relies on five core libraries from Python and R, each playing a
critical role in the analysis pipeline. In Python, \texttt{pillow}
\citep{clark2015pillow} handles image processing tasks such as reading
and resizing PNG files of residual plots, then converting them into
input tensors for further analysis. \texttt{TensorFlow}
\citep{abadi2016tensorflow}, a key component of modern machine learning,
is used to predict the VSS of these plots using a pre-trained
convolutional neural network.

In the R environment, \texttt{autovi} utilizes several libraries.
\texttt{ggplot2} \citep{ggplot2} generates the initial residual plots,
saved as PNG files for visual input. \texttt{cassowaryr}
\citep{mason2022cassowaryr} computes scagnostics (scatter plot
diagnostics), providing numerical features that capture statistical
properties of the plots. These scagnostics complement the visual
analysis by offering quantitative metrics as secondary input to the
computer vision model. \texttt{reticulate} \citep{reticulate} enables
seamless communication between R and Python.

@@

The \texttt{autovi} package is available on CRAN. It is actively
developed and maintained, with the latest updates accessible on GitHub.
This paper uses \texttt{autovi} version 0.4.2. The package includes
internal functions to check the current Python environment used by the
\texttt{reticulate} package. If the necessary Python packages are not
installed in the Python interpreter, an error will be raised. If you
want to select a specific Python environment, you can do so by calling
the \texttt{reticulate::use\_python()} function before using the
\texttt{autovi} package.

We recommend using the Shiny app \texttt{autovi.web} if users encounter
installation problems.

@@

\caption{\label{fig-autovi-diag}Diagram illustrating the infrastructure
of the R package \texttt{autovi}. The modules in green are primary
inputs provided by users. Modules in blue are overridable methods that
can be modified to accommodate users' specific needs. The module in
yellow is a pre-defined non-overridable method. The modules in red are
primary outputs of the package.}

@@

The \texttt{autovi} R package can be extended to accommodate other
classes of models beyond linear regression, such as generalized linear
models (\texttt{glm}). This is achieved by substituting the relevant
overridable modules, and if needed, supplying a different \texttt{Keras}
model.

@@

The \texttt{autovi.web} shiny application extends the functionality of
\texttt{autovi} by offering a user-friendly web interface for automated
residual plot assessment. This eliminates the common challenges
associated with software installation, so users can avoid managing
Python environments or handling version requirements for R libraries.
The platform is cross-platform and accessible on various devices and
operating systems, making it suitable even for users without R
programming experience. Additionally, updates are managed centrally,
ensuring that users always have access to the latest features. This
section discusses the implementation based on \texttt{autovi.web}
version 0.1.0.

@@

The interface \texttt{autovi.web} is built using the \texttt{shiny}
\citep{shiny} and \texttt{shinydashboard} \citep{shinydashboard} R
packages. Hosted on the \href{https://www.shinyapps.io}{shinyapps.io}
domain, the application is accessible through any modern web browser.
The R packages \texttt{htmltools} \citep{htmltools} and
\texttt{shinycssloaders} \citep{shinycssloaders} are used to render
markdown documentation in shiny application, and for loading animations
for shiny widgets, respectively.

@@

We also explored the possibility of implementing the web interface using
frameworks built on other languages, such as Python. However, server
hosting domains that natively support Python servers typically do not
have the latest version of R installed. Additionally, calling R from
Python is typically done using the \texttt{rpy2} Python library
\citep{rpy2}, but this approach can be awkward when dealing with
language syntax related to non-standard evaluation. Another option we
considered was renting a server where we could have full control, such
as those provided by cloud platforms like Google Cloud Platform (GCP) or
Amazon Web Services (AWS). However, deploying and maintaining the server
securely requires some expertise. Ultimately, the most practical
solution was to use the \texttt{shiny} and \texttt{shinydashboard}
frameworks, which are well-established in the R community and offer a
solid foundation for web application development.

The server-side configuration of \texttt{autovi.web} is carefully
designed to support its functionality. Most required Python libraries,
including \texttt{pillow} and \texttt{numpy}, are pre-installed on the
server. These libraries are integrated into the Shiny application using
the \texttt{reticulate} package, which provides an interface between R
and Python.

Due to shinyapps.io's resource policy, inactive servers enter sleep
mode, clearing the local Python environment. When reactivated for a new
session, libraries must be reinstalled. While this ensures a clean
environment for each session, it may lead to slightly longer loading
times for the first user after a period of inactivity.

@@

While \texttt{autovi} requires downloading the pre-trained computer
vision models from GitHub, these models in ``.keras'' file format are
incompatible with \texttt{TensorFlow.js}. Therefore, we extract and
store the model weights in JSON files and include them as extra
resources in the Shiny application. When the application initializes,
\texttt{TensorFlow.js} rebuilds the computer vision model using these
pre-stored weights.

To allow communication between \texttt{TensorFlow.js} and other
components of the Shiny application, the \texttt{shinyjs} R package
\citep{shinyjs} is used. This package allows calling custom JavaScript
code within the Shiny framework. The specialized JavaScript code for
initializing \texttt{TensorFlow.js} and calling \texttt{TensorFlow.js}
for VSS prediction is deployed alongside the Shiny application as
additional resources.

@@

The workflow of \texttt{autovi.web} is designed to be straightforward,
with numbered steps displayed in each panel. There are two example
datasets provided by the web application. The single residual plot
example uses the \texttt{dino} dataset from the R package
\texttt{datasauRus} \citep{datasaurus}. The lineup example uses
residuals from a simulated regression model that has a non-linearity
issue. We walk through the lineup example to further demonstrate the
workflow of the web application.

@@

This paper presents new regression diagnostics software, the R package
\texttt{autovi} and its accompanying web interface, \texttt{autovi.web}.
It addresses a critical gap in the current landscape of statistical
software. While regression tools are widely available, effective and
efficient diagnostic methods have lagged behind, particularly in the
field of residual plot interpretation.

The \texttt{autovi} R package, introduced in this paper, automates the
assessment of residual plots by incorporating a computer vision model,
reducing reliance on time-consuming and potentially inconsistent human
interpretation. This automation improves the efficiency of the
diagnostic process and promotes consistency in model evaluation across
different users and studies.

The development of the accompanying Shiny app, \texttt{autovi.web},
expands access to these advanced diagnostic tools, by providing a
user-friendly interface. It makes automated residual plot assessment
accessible to a broader audience, including those who may not have
extensive programming experience. This web-based solution effectively
addresses the potential barriers to adoption, such as complex
dependencies and installation requirements, that are often associated
with advanced statistical software.

@@

The current version of \texttt{autovi} can be installed from CRAN, and
source code for both packages are available at
\href{https://github.com/TengMCing/autovi}{\texttt{github.com/TengMCing/autovi}}
and
\href{https://github.com/TengMCing/autovi_web}{\texttt{github.com/TengMCing/autovi\_web}}
respectively. The web interface is available from
\href{https://autoviweb.netlify.app/}{\texttt{autoviweb.netlify.app}}.

@@

These \texttt{R} packages were used for the work: \texttt{tidyverse}
\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{kableExtra}
\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
\texttt{rcartocolor} \citep{rcartocolor}, \texttt{glue} \citep{glue},
\texttt{here} \citep{here}, \texttt{magick} \citep{magick},
\texttt{yardstick} \citep{yardstick} and \texttt{reticulate}
\citep{reticulate}.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Automated residual plot assessment with the  package  and the  app },
  pdfauthor={Weihao Li; Dianne Cook; Emi Tanaka; Susan VanderPlas; Klaus Ackermann},
  pdfkeywords={initial data analysis, statistical graphics, data
visualization, visual inference, computer vision, machine
learning, hypothesis testing, regression analysis, model diagnostics},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{Automated residual plot assessment with the \textsf{R} package
\textsf{autovi} and the \textsf{Shiny} app \textsf{autovi.web}}

@@

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\hat{\boldsymbol{e}})}{q(\hat{\boldsymbol{e}})}p(\hat{\boldsymbol{e}})d\hat{\boldsymbol{e}},
\end{equation}

here, \(\hat{\boldsymbol{e}}\) denotes the residual vector from the
regression model, and \(p(\cdot)\) and \(q(\cdot)\) are the probability
density functions of the reference residual distribution \(\mathrm{P}\)
and the true residual distribution \(\mathrm{Q}\), respectively.

This distance measure depends on knowledge of the true residual
distribution, which is unknown in practice. To compute \(D_{KL}\) for
the training samples, Equation (\ref{eq:kl-1}) takes different forms
depending on the specific model violations. For instance, where
necessary higher-order predictors, \(\boldsymbol{Z}\), and their
corresponding parameter, \(\boldsymbol{\beta}_Z\), are omitted from the
fitted linear model, the distance measure can be expanded as follows:

@@

\section{\texorpdfstring{\textsf{R} package:
autovi}{ package: autovi}}\label{sec-autovi}

The main purpose of \texttt{autovi} is to provide rejection decisions
and \(p\)-values for testing the null hypothesis (\(H_0\)) that the
regression model is correctly specified. The package provides automated
interpretation of residual plots using computer vision. The name
\texttt{autovi} stands for \textbf{auto}mated \textbf{v}isual
\textbf{i}nference. This functionality can be accessed through the
\textsf{R} package \texttt{autovi}, or through a web interface,
\texttt{autovi.web}, which enables users to perform analyses without
installing \textsf{R}, \textsf{Python}, or their associated dependencies
locally.

@@

The \texttt{autovi} infrastructure effectively integrates multiple
programming languages and libraries into a comprehensive analytical
tool. It relies on five core libraries from \textsf{Python} and
\textsf{R}, each playing a critical role in the analysis pipeline. In
\textsf{Python}, \texttt{pillow} \citep{clark2015pillow} handles image
processing tasks such as reading and resizing PNG files of residual
plots, then converting them into input tensors for further analysis.
\texttt{TensorFlow} \citep{abadi2016tensorflow}, a key component of
modern machine learning, is used to predict the VSS of these plots using
a pre-trained convolutional neural network.

In the \textsf{R} environment, \texttt{autovi} utilizes several
libraries. \texttt{ggplot2} \citep{ggplot2} generates the initial
residual plots, saved as PNG files for visual input. \texttt{cassowaryr}
\citep{mason2022cassowaryr} computes scagnostics (scatter plot
diagnostics), providing numerical features that capture statistical
properties of the plots. These scagnostics complement the visual
analysis by offering quantitative metrics as secondary input to the
computer vision model. \texttt{reticulate} \citep{reticulate} enables
seamless communication between \textsf{R} and \textsf{Python}.

@@

The \texttt{autovi} package is available on CRAN. It is actively
developed and maintained, with the latest updates accessible on GitHub.
This paper uses \texttt{autovi} version 0.4.2. The package includes
internal functions to check the current \textsf{Python} environment used
by the \texttt{reticulate} package. If the necessary \textsf{Python}
packages are not installed in the \textsf{Python} interpreter, an error
will be raised. If you want to select a specific \textsf{Python}
environment, you can do so by calling the
\texttt{reticulate::use\_python()} function before using the
\texttt{autovi} package.

We recommend using the \textsf{Shiny} app \texttt{autovi.web} if users
encounter installation problems.

@@

\caption{\label{fig-autovi-diag}Diagram illustrating the infrastructure
of the extsf\{R\} package \texttt{autovi}. The modules in green are
primary inputs provided by users. Modules in blue are overridable
methods that can be modified to accommodate users' specific needs. The
module in yellow is a pre-defined non-overridable method. The modules in
red are primary outputs of the package.}

@@

The \texttt{autovi} \textsf{R} package can be extended to accommodate
other classes of models beyond linear regression, such as generalized
linear models (\texttt{glm}). This is achieved by substituting the
relevant overridable modules, and if needed, supplying a different
\texttt{Keras} model.

@@

The \texttt{autovi.web} \textsf{Shiny} application extends the
functionality of \texttt{autovi} by offering a user-friendly web
interface for automated residual plot assessment. This eliminates the
common challenges associated with software installation, so users can
avoid managing \textsf{Python} environments or handling version
requirements for \textsf{R} libraries. The platform is cross-platform
and accessible on various devices and operating systems, making it
suitable even for users without \textsf{R} programming experience.
Additionally, updates are managed centrally, ensuring that users always
have access to the latest features. This section discusses the
implementation based on \texttt{autovi.web} version 0.1.0.

@@

The interface \texttt{autovi.web} is built using the \texttt{shiny}
\citep{shiny} and \texttt{shinydashboard} \citep{shinydashboard}
\textsf{R} packages. Hosted on the
\href{https://www.shinyapps.io}{shinyapps.io} domain, the application is
accessible through any modern web browser. The \textsf{R} packages
\texttt{htmltools} \citep{htmltools} and \texttt{shinycssloaders}
\citep{shinycssloaders} are used to render markdown documentation in
\textsf{Shiny} application, and for loading animations for
\textsf{Shiny} widgets, respectively.

@@

We also explored the possibility of implementing the web interface using
frameworks built on other languages, such as \textsf{Python}. However,
server hosting domains that natively support \textsf{Python} servers
typically do not have the latest version of \textsf{R} installed.
Additionally, calling \textsf{R} from \textsf{Python} is typically done
using the \texttt{rpy2} \textsf{Python} library \citep{rpy2}, but this
approach can be awkward when dealing with language syntax related to
non-standard evaluation. Another option we considered was renting a
server where we could have full control, such as those provided by cloud
platforms like Google Cloud Platform (GCP) or Amazon Web Services (AWS).
However, deploying and maintaining the server securely requires some
expertise. Ultimately, the most practical solution was to use the
\texttt{shiny} and \texttt{shinydashboard} frameworks, which are
well-established in the \textsf{R} community and offer a solid
foundation for web application development.

The server-side configuration of \texttt{autovi.web} is carefully
designed to support its functionality. Most required \textsf{Python}
libraries, including \texttt{pillow} and \texttt{numpy}, are
pre-installed on the server. These libraries are integrated into the
\textsf{Shiny} application using the \texttt{reticulate} package, which
provides an interface between \textsf{R} and \textsf{Python}.

Due to shinyapps.io's resource policy, inactive servers enter sleep
mode, clearing the local \textsf{Python} environment. When reactivated
for a new session, libraries must be reinstalled. While this ensures a
clean environment for each session, it may lead to slightly longer
loading times for the first user after a period of inactivity.

@@

While \texttt{autovi} requires downloading the pre-trained computer
vision models from GitHub, these models in ``.keras'' file format are
incompatible with \texttt{TensorFlow.js}. Therefore, we extract and
store the model weights in JSON files and include them as extra
resources in the \textsf{Shiny} application. When the application
initializes, \texttt{TensorFlow.js} rebuilds the computer vision model
using these pre-stored weights.

To allow communication between \texttt{TensorFlow.js} and other
components of the \textsf{Shiny} application, the \texttt{shinyjs}
\textsf{R} package \citep{shinyjs} is used. This package allows calling
custom JavaScript code within the \textsf{Shiny} framework. The
specialized JavaScript code for initializing \texttt{TensorFlow.js} and
calling \texttt{TensorFlow.js} for VSS prediction is deployed alongside
the \textsf{Shiny} application as additional resources.

@@

The workflow of \texttt{autovi.web} is designed to be straightforward,
with numbered steps displayed in each panel. There are two example
datasets provided by the web application. The single residual plot
example uses the \texttt{dino} dataset from the \textsf{R} package
\texttt{datasauRus} \citep{datasaurus}. The lineup example uses
residuals from a simulated regression model that has a non-linearity
issue. We walk through the lineup example to further demonstrate the
workflow of the web application.

@@

This paper presents new regression diagnostics software, the \textsf{R}
package \texttt{autovi} and its accompanying web interface,
\texttt{autovi.web}. It addresses a critical gap in the current
landscape of statistical software. While regression tools are widely
available, effective and efficient diagnostic methods have lagged
behind, particularly in the field of residual plot interpretation.

The \texttt{autovi} \textsf{R} package, introduced in this paper,
automates the assessment of residual plots by incorporating a computer
vision model, reducing reliance on time-consuming and potentially
inconsistent human interpretation. This automation improves the
efficiency of the diagnostic process and promotes consistency in model
evaluation across different users and studies.

The development of the accompanying \textsf{Shiny} app,
\texttt{autovi.web}, expands access to these advanced diagnostic tools,
by providing a user-friendly interface. It makes automated residual plot
assessment accessible to a broader audience, including those who may not
have extensive programming experience. This web-based solution
effectively addresses the potential barriers to adoption, such as
complex dependencies and installation requirements, that are often
associated with advanced statistical software.

@@

The current version of \textsf{autovi} can be installed from CRAN, and
source code for both packages are available at
\href{https://github.com/TengMCing/autovi}{\texttt{github.com/TengMCing/autovi}}
and
\href{https://github.com/TengMCing/autovi_web}{\texttt{github.com/TengMCing/autovi\_web}}
respectively. The web interface is available from
\href{https://autoviweb.netlify.app/}{\texttt{autoviweb.netlify.app}}.

@@

These \textsf{R} packages were used for the work: \textsf{tidyverse}
\citep{tidyverse}, \textsf{lmtest} \citep{lmtest}, \textsf{kableExtra}
\citep{kableextra}, \textsf{patchwork} \citep{patchwork},
\textsf{rcartocolor} \citep{rcartocolor}, \textsf{glue} \citep{glue},
\textsf{here} \citep{here}, \textsf{magick} \citep{magick},
\textsf{yardstick} \citep{yardstick} and \textsf{reticulate}
\citep{reticulate}.
